\texttt{MSnbase}, efficient R-based access and manipulation of raw mass spectrometry data

### 2020-04-25 msnbase2.Rnw
replace bench1 tab with figure

<<setup, include = FALSE, message = FALSE>>=
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
library("xtable")
library("microbenchmark")
library("tidyverse")
@

@@

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load
metadata in memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Table \ref{tab:bench1} shows 4-fold faster reading times
(left) and 10-fold smaller memory footprint (right).

<<time_sz>>=
load("bench_time_sz.rda")
time <- sapply(time_sz, function(x) x[, "time"])
sz <- sapply(time_sz, function(x) x[, "sz"])
colnames(sz) <- colnames(time) <- c(1, 5, 10)
rownames(sz) <- rownames(time) <- c("in-memory", "on-disk")
sz <- round(sz/(1014^2), 2)
print(xtable(time), floating = FALSE, file = "tabs/t_read.tex")
print(xtable(sz), floating = FALSE, include.rownames = FALSE, file = "tabs/sz.tex")
@

\begin{table}[ht]
\centering
\subfloat[Reading time (s)]{\label{tab:tread}\scalebox{1}{\input{./tabs/t_read}}}\quad
\subfloat[Object size (MB)]{\label{tab:sz}\scalebox{1}{\input{./tabs/sz}}}
\caption{Reading time (in seconds) (left) and data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. }
\label{tab:bench1}
\end{table}

@@

As shown on figure \ref{fig:filt}, this lazy mechanism si
significantly faster than its application on in-memory data.

@@

The advantageous reading and execution times and memory footprint of
the on-disk backend are possible by avoiding unnecessary access to the
raw data. Once access to the spectra M/Z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on figure
\ref{fig:access}. This gain is maximal when the whole dataset is the
be accessed (i.e. all spectra are already in memory) and negligeable
when large fractions of the data need to be subset.

@@

Enables large scale data analyses (metabolomics example, see Johannes)

>>>>

<<setup, include = FALSE, message = FALSE>>=
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
library("xtable")
library("microbenchmark")
library("tidyverse")
library("patchwork")
@

@@

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load metadata in
memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Figure \ref{fig:bench1} shows 4-fold faster reading times
(left) and 10-fold smaller memory footprint (right).

<<time_sz>>=
load("bench_time_sz.rda")
time <- sapply(time_sz, function(x) x[, "time"])
sz <- sapply(time_sz, function(x) x[, "sz"])
colnames(sz) <- colnames(time) <- c(1, 5, 10)
rownames(sz) <- rownames(time) <- c("in-memory", "on-disk")
sz <- round(sz/(1014^2), 2)

p1 <- data.frame(time) %>% 
    rownames_to_column() %>% 
    pivot_longer(names_to = "files", 
                 values_to = "time", 
                 -rowname) %>% 
    rename(backend = rowname) %>% 
    mutate(files = as.numeric(sub("X", "", files))) %>% 
    ggplot(aes(x = files, y = time, colour = backend)) + 
    geom_point() + 
    geom_line() + 
    scale_x_continuous(breaks=c(1, 5, 10)) + 
    ylab("Reading time (s)") + 
    theme(legend.position = "none")

p2 <- data.frame(sz) %>% 
    rownames_to_column() %>% 
    pivot_longer(names_to = "files", 
                 values_to = "size",
                 -rowname) %>% 
    rename(backend = rowname) %>% 
    mutate(files = as.numeric(sub("X", "", files))) %>% 
    ggplot(aes(x = files, y = size, colour = backend)) + 
    geom_point() + 
    geom_line() + 
    scale_x_continuous(breaks=c(1, 5, 10)) + 
    ylab("Object size (MB)") 
@

\begin{figure}[ht]
  \centering
<<plot_bench1, out.width = '100%', fig.width = 7, fig.height = 3>>=
p1 + p2 + plot_annotation(tag_levels = 'a')
@
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. }
\label{fig:bench1}
\end{figure}

@@

As shown on figure~\ref{fig:filt}, this lazy mechanism si
significantly faster than its application on in-memory data.

@@

The advantageous reading and execution times and memory footprint of
the on-disk backend are possible by avoiding unnecessary access to the
raw data. Once access to the spectra M/Z and intensity values become
mandatory (for example for plotting), then the in-memory backend
becomes more efficient, as illustrated on
figure~\ref{fig:access}. This gain is maximal when the whole dataset
is the be accessed (i.e. all spectra are already in memory) and
negligeable when large fractions of the data need to be subset.

@@

This new on-disk infrastructure enables large scale data analyses
using \texttt{MSnbase} (metabolomics example, see Johannes).

### 2020-04-25 msnbase2.tex
replace bench1 tab with figure

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load
metadata in memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Table \ref{tab:bench1} shows 4-fold faster reading times
(left) and 10-fold smaller memory footprint (right).

\begin{table}[ht]
\centering
\subfloat[Reading time (s)]{\label{tab:tread}\scalebox{1}{\input{./tabs/t_read}}}\quad
\subfloat[Object size (MB)]{\label{tab:sz}\scalebox{1}{\input{./tabs/sz}}}
\caption{Reading time (in seconds) (left) and data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. }
\label{tab:bench1}
\end{table}

>>>>

The former is the legacy storage mode, implemented in the first
version of the package, that loads all the raw data and the metadata
in memory. This solution doesn't scale for modern large dataset, and
was complemented by the on-disk backend, that only load metadata in
memory and accesses the spectra in the original files when
needed. There are two direct benefits using the on-disk
backend. Figure \ref{fig:bench1} shows 4-fold faster reading times
(left) and 10-fold smaller memory footprint (right).

\begin{figure}[ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=1\linewidth]{figure/plot_bench1-1}

\end{knitrout}
\caption{(a) Reading time (in seconds) and (b) data size in memory (in
  MB) to read/store 1, 5 and 10 files containing 1431 MS1 (on-disk
  only) and 6103 MS2 (on-disk and in-memory) spectra. }
\label{fig:bench1}
\end{figure}
